---
title: "Coursera Fitabase Data Case Study"
author: "Joseph Lennan"
date: "2022-09-28"
output: html_document
---

```{r Notepad, eval=FALSE, include=FALSE}

# To Do -------------------------------------------------------------------

#Possible Future Analysis
  #Calories Hour(s) before bed - Affect time to go to sleep/time in bed?
  #Hourly Intensities with Sleep

```

# Explanation

The goal of this case study for the Google Analytics Certificate is to analyze the provided smart fitness data to provide insight into trends and potential opportunities with Bellabeat's current products in mind.

# Documentation

Documentation of Data
  https://www.fitabase.com/media/1930/fitabasedatadictionary102320.pdf
  Distance in Kilometers
  Calories is Calories burned


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(skimr)
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)
```

#Data Import & Prep

```{r import, include=FALSE}

#File Storage for Data
setwd("C:/Users/lenna/Downloads/Coding/R Studio WD/Coursera_Case_Study_Fitabase Data 4.12.16-5.12.16")

#CSV List in Working Directory
file_list = list.files(pattern="*.csv")

#Read CSVs in Working Directory
myfiles = lapply(file_list, read.csv,header=TRUE,sep=",",quote="")

#Set myfiles Name to Name In Folder
names(myfiles) = file_list
```

```{r Data Prep}
#Get Dataframes

sleepDay = myfiles$sleepDay_merged.csv
dailyActivity = myfiles$dailyActivity_merged.csv
dailyIntensities = myfiles$dailyIntensities_merged.csv
dailySteps = myfiles$dailySteps_merged.csv
dailyCalories = myfiles$dailyCalories_merged.csv
hourlyCalories = myfiles$hourlyCalories_merged.csv
hourlyIntensities = myfiles$hourlyIntensities_merged.csv
weightLog = myfiles$weightLogInfo_merged.csv

```

#Data Examination

```{r Data Skimming}
#Skim Data for key observations & potential flaws

sleepDay %>%
  select(TotalSleepRecords,
         TotalMinutesAsleep,
         TotalTimeInBed
  ) %>%
  skim_without_charts()
#Mean 419 Min Asleep, 39.2 Mean Extra Minutes in Bed
#Standard Deviations seem high, ~68% of people take ~86 minutes to go to bed?
#Might be counting time after waking and still lying in bed?



dailyActivity %>% 
  select(
    TotalDistance,
    TotalSteps,
    Calories,
    TrackerDistance,
    LoggedActivitiesDistance,
    SedentaryActiveDistance,
    SedentaryMinutes,
    LightActiveDistance,
    LightlyActiveMinutes,
    ModeratelyActiveDistance,
    FairlyActiveMinutes,
    VeryActiveDistance,
    VeryActiveMinutes
  ) %>% 
  skim_without_charts()
#5.49 Units Mean Total Distance, 7638 Mean Steps, 2304 Mean Calories...

dailyIntensities %>% 
  select(
    SedentaryMinutes,
    LightlyActiveMinutes,
    FairlyActiveMinutes,
    VeryActiveMinutes,
    SedentaryActiveDistance,
    LightActiveDistance,
    ModeratelyActiveDistance,
    VeryActiveDistance
  ) %>% 
  skim_without_charts()
#193 mean lightly active minutes, 32 fairly active mean minutes


dailySteps %>% 
  select(
    StepTotal
  ) %>% 
  skim_without_charts()

dailyCalories %>% 
  select(
    Calories
  ) %>% 
  skim_without_charts()

```

```{r Data Editing}

#Separate Datetime to Date and Time

#Date Time - Posix*
sleepDay$Date_Time_Cleaned = strptime(sleepDay$SleepDay,"%m/%d/%Y %I:%M:%S %p",tz="UTC")
hourlyCalories$Date_Time_Cleaned = strptime(hourlyCalories$ActivityHour,"%m/%d/%Y %I:%M:%S %p",tz="UTC")
hourlyIntensities$Date_Time_Cleaned = strptime(hourlyIntensities$ActivityHour,"%m/%d/%Y %I:%M:%S %p",tz="UTC")
weightLog$Date_Time_Cleaned = strptime(weightLog$Date,"%m/%d/%Y %I:%M:%S %p",tz="UTC")


#Date Only - Posix*
sleepDay$Date_Cleaned = strptime(sleepDay$SleepDay,"%m/%d/%Y")
hourlyCalories$Date_Cleaned = strptime(hourlyCalories$ActivityHour,"%m/%d/%Y")
hourlyIntensities$Date_Cleaned = strptime(hourlyIntensities$ActivityHour,"%m/%d/%Y")
weightLog$Date_Cleaned = strptime(weightLog$Date,"%m/%d/%Y")

dailyActivity$Date_Cleaned = strptime(dailyActivity$ActivityDate,"%m/%d/%Y")
dailyIntensities$Date_Cleaned = strptime(dailyIntensities$ActivityDay,"%m/%d/%Y")
dailySteps$Date_Cleaned = strptime(dailySteps$ActivityDay,"%m/%d/%Y")
dailyCalories$Date_Cleaned = strptime(dailyCalories$ActivityDay,"%m/%d/%Y")

#Time Only - Characters
sleepDay$Time_Cleaned = strftime(sleepDay$Date_Time_Cleaned,"%H:%M:%S")
hourlyCalories$Time_Cleaned = strftime(hourlyCalories$Date_Time_Cleaned,"%H:%M:%S")
hourlyIntensities$Time_Cleaned = strftime(hourlyIntensities$Date_Time_Cleaned,"%H:%M:%S")
weightLog$Time_Cleaned = strftime(weightLog$Date_Time_Cleaned,"%H:%M:%S")


#Joins/Merge

#Activity and Sleep
Activity_Day_Sleep_Day = left_join(dailyActivity,sleepDay,by=c("Id","Date_Cleaned"))


#Adding Columns for Analysis

#Time Awake in Bed
Activity_Day_Sleep_Day$SleepLatency = Activity_Day_Sleep_Day$TotalTimeInBed - Activity_Day_Sleep_Day$TotalMinutesAsleep

#Conversion Integer to Numeric - For Correlations
Activity_Day_Sleep_Day[c(3,4,5,6,7,8,9,10,11,12,13,14,15,19,20,23)] = sapply(Activity_Day_Sleep_Day[c(3,4,5,6,7,8,9,10,11,12,13,14,15,19,20,23)],as.numeric)



```

```{r Data Graphing and Observation, message=FALSE, warning=FALSE}

# Examining & Graphing Variables


#Sleep Variable
Activity_Day_Sleep_Day %>%
  select(TotalSleepRecords,
         TotalMinutesAsleep,
         TotalTimeInBed,
         SleepLatency
  ) %>%
  skim_without_charts()

#Individual Means and Deviations - Some Notable Outliers
#Some IDs have no SD - Because only one observed sleep data. 
#Notably, two individuals had substantially higher awake means, 
#2 others had substantially higher awake standard deviations

Sleep_Awake_Mean_SD = 
  Activity_Day_Sleep_Day %>%
  group_by(Id) %>%
  drop_na() %>%
  mutate(awake_mean=mean(SleepLatency,na.rm=TRUE),
         awake_sd = sd(SleepLatency,na.rm=TRUE),
         asleep_mean=mean(TotalMinutesAsleep,na.rm=TRUE),
         asleep_sd = sd(TotalMinutesAsleep,TRUE),
         .keep="used") %>%
  filter(!duplicated(Id))

#Visualization of Outliers around cluster of more normal means and deviations
ggplot(Sleep_Awake_Mean_SD,aes(x=asleep_mean,y=asleep_sd)) + geom_point()

ggplot(Sleep_Awake_Mean_SD,aes(x=awake_mean,y=awake_sd)) + geom_point()

#Exercise/Activity Variable
Exercise_Sedentary_Mean_SD = 
  Activity_Day_Sleep_Day %>%
  group_by(Id) %>%
  drop_na() %>%
  mutate(TotalSteps_mean=mean(TotalSteps,na.rm=TRUE),
         TotalSteps_sd = sd(TotalSteps,na.rm=TRUE),
         TotalDistance_mean=mean(TotalDistance,na.rm=TRUE),
         TotalDistance_sd = sd(TotalDistance,TRUE),
         Calories_mean=mean(Calories,na.rm=TRUE),
         Calories_sd = sd(Calories,na.rm=TRUE),
         SedentaryMinutes_mean=mean(SedentaryMinutes,na.rm=TRUE),
         SedentaryMinutes_sd = sd(SedentaryMinutes,TRUE),
         .keep="used") %>%
  filter(!duplicated(Id))

ggplot(Exercise_Sedentary_Mean_SD,aes(x=TotalSteps_mean,y=TotalSteps_sd)) + geom_point()

ggplot(Exercise_Sedentary_Mean_SD,aes(x=TotalDistance_mean,y=TotalDistance_sd)) + geom_point()

ggplot(Exercise_Sedentary_Mean_SD,aes(x=Calories_mean,y=Calories_sd)) + geom_point()

ggplot(Exercise_Sedentary_Mean_SD,aes(x=SedentaryMinutes_mean,y=SedentaryMinutes_sd)) + geom_point()


#Minutes - VeryActiveMinutes, FairlyActiveMinutes,LightlyActiveMinutes,SedentaryMinutes

Exercise_Minutes_Mean_SD = 
  Activity_Day_Sleep_Day %>%
  group_by(Id) %>%
  drop_na() %>%
  mutate(VeryActiveMinutes_mean=mean(VeryActiveMinutes,na.rm=TRUE),
         VeryActiveMinutes_sd = sd(VeryActiveMinutes,na.rm=TRUE),
         FairlyActiveMinutes_mean=mean(FairlyActiveMinutes,na.rm=TRUE),
         FairlyActiveMinutes_sd = sd(FairlyActiveMinutes,TRUE),
         LightlyActiveMinutes_mean=mean(LightlyActiveMinutes,na.rm=TRUE),
         LightlyActiveMinutes_sd = sd(LightlyActiveMinutes,na.rm=TRUE),
         SedentaryMinutes_mean=mean(SedentaryMinutes,na.rm=TRUE),
         SedentaryMinutes_sd = sd(SedentaryMinutes,TRUE),
         .keep="used") %>%
  filter(!duplicated(Id))

ggplot(Exercise_Minutes_Mean_SD,aes(x=VeryActiveMinutes_mean,y=VeryActiveMinutes_sd)) + geom_point()

ggplot(Exercise_Minutes_Mean_SD,aes(x=FairlyActiveMinutes_mean,y=FairlyActiveMinutes_sd)) + geom_point()

ggplot(Exercise_Minutes_Mean_SD,aes(x=LightlyActiveMinutes_mean,y=LightlyActiveMinutes_sd)) + geom_point()

ggplot(Exercise_Minutes_Mean_SD,aes(x=SedentaryMinutes_mean,y=SedentaryMinutes_sd)) + geom_point()


#Graphs

#Basic Graphs from case study description

ggplot(data=Activity_Day_Sleep_Day,aes(x=TotalSteps,y=SedentaryMinutes)) + geom_point() + geom_smooth()

ggplot(data=Activity_Day_Sleep_Day,aes(x=TotalMinutesAsleep,y=TotalTimeInBed)) + geom_point() + geom_smooth()

```





# Graphs and Correlation

```{r Important/Correlation Graphs}

#Investigating Correlations alongside Graphs

#High Correlation between Steps and Calories Burned
ggplot(data=Activity_Day_Sleep_Day,aes(x=TotalSteps,y=Calories)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$TotalSteps,Activity_Day_Sleep_Day$Calories,use="complete.obs")

#High Correlation - Less Sleep - More Sedentary Minutes, but also some outliers with 0 Sedentary Minutes
ggplot(data=Activity_Day_Sleep_Day, aes(x=SedentaryMinutes, y=TotalMinutesAsleep)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$SedentaryMinutes,Activity_Day_Sleep_Day$TotalMinutesAsleep,use="complete.obs")

#High (Obvious) Correlation More Distance = More Calories Burned
ggplot(data=Activity_Day_Sleep_Day, aes(x=TotalDistance, y=Calories)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$TotalDistance,Activity_Day_Sleep_Day$Calories,use="complete.obs")


#Low Negative Correlation - Sleep Latency might be caused by not enough calories burned for excess energy?
ggplot(data=Activity_Day_Sleep_Day, aes(x=Calories, y=SleepLatency)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$Calories,Activity_Day_Sleep_Day$SleepLatency,use="complete.obs")
  #What if we look at the range from 1000-2000 Calories?
  Activity_Day_Sleep_Day %>%
    filter(between(Calories,1000,2000)) %>%
    ggplot(aes(x=Calories,y=SleepLatency)) + geom_point() + geom_smooth()
  Activity_Day_Sleep_Day %>%
    filter(between(Calories,1000,2000)) %>%
    summarise(cor(Calories, SleepLatency, use = "complete.obs"))

#More Exercise and Less Sedentary Minutes = Less Minutes Asleep (Possibly more rested or better quality sleep?)
ggplot(data=Activity_Day_Sleep_Day, aes(x=TotalDistance, y=TotalMinutesAsleep)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$TotalDistance,Activity_Day_Sleep_Day$TotalMinutesAsleep,use="complete.obs")

ggplot(data=Activity_Day_Sleep_Day, aes(x=TotalDistance, y=TotalTimeInBed)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$TotalDistance,Activity_Day_Sleep_Day$TotalTimeInBed,use="complete.obs")


```


```{r Low-Value Data}
#Low Correlation Data

#Calories Burned vs Sleep - 
#Maybe having too much energy from not burning enough for some people?
#No clear relationship between Minutes Asleep and Calories burned here
ggplot(data=Activity_Day_Sleep_Day, aes(x=Calories, y=TotalMinutesAsleep)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$Calories,Activity_Day_Sleep_Day$TotalMinutesAsleep,use="complete.obs")


#Normal Minutes Asleep (~5-8 hrs), people that sleep less more likely to spend more time in bed (trying to go back to sleep)
ggplot(data=Activity_Day_Sleep_Day, aes(x=SleepLatency, y=TotalMinutesAsleep)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$SleepLatency,Activity_Day_Sleep_Day$TotalMinutesAsleep,use="complete.obs")


#Calories and Total Time in Bed
ggplot(data=Activity_Day_Sleep_Day,aes(x=Calories,y=TotalTimeInBed)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$Calories,Activity_Day_Sleep_Day$TotalTimeInBed,use="complete.obs")


#Does More Exercise Lower Sleep Latency? Not such a clear relationship, possibly too many external variables, possibly misunderstanding TimeAwakeInBed Variable
ggplot(data=Activity_Day_Sleep_Day, aes(x=TotalDistance, y=SleepLatency)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$Calories,Activity_Day_Sleep_Day$TotalTimeInBed,use="complete.obs")

ggplot(data=Activity_Day_Sleep_Day, aes(x=SedentaryMinutes, y=SleepLatency)) + geom_point() + geom_smooth()
cor(Activity_Day_Sleep_Day$Calories,Activity_Day_Sleep_Day$TotalTimeInBed,use="complete.obs")



```

# Data Bulletpoints

More Distance and Steps Correlated to more Calories Burned
Less Sleep Correlated with More Sedentary Minutes that day
  Less sleep, lower energy for exercise?
More Exercise and Less Sedentary Minutes = Less Minutes Asleep
  More exercise, higher sleep quality and thus less minutes asleep? 
Low Negative Correlation between Calories and Sleep Latency
  Could this be due to less calories burned keeping people awake? Lack of Food or Exercise?



# Product Suggestions

My product suggestion or improvement for Bellabeat would be directed at aiming to improve sleep and exercise, working to balance and help schedule Bellabeat users both exercise consistently and sleep consistently for better motivation and higher quality sleep.

By further analyzing what general data says about the quality and quantity of sleep in correlation with various factors of exercise, such as the time exercise takes place as well as the length and intensity of it, we can draw some ideas on what general suggestions to make to people. 

Depending on the future analysis, we might break down users by their fitness and make suggestions such as getting at least 30 minutes of intense exercise per day and encourage them to consistently sleep for healthy amounts to help them reinforce their exercise habits with the extra willpower and motivation that can come from it. 

I'm not an expert when it comes to mobile apps and so I think it would be best to explore how mobile apps encourage users' habits utilizing rewards and other methods to see what would work best for Bellabeat and their users. One small way I can think of to encourage users is that once they have been active enough, showing off the data we've tracked about their progress and success, such as streaks or improvements in exercise and sleep.

Figuring out how to implement these app improvements would be highly beneficial to Bellabeat and their users by helping reinforce and benefit exercise and sleep, and thereby the use of the app as a means to do so.

